Symmetric Circuits for Rank Gates

### 2018-03-04 sections/symmetric-circuits.tex
I have made changes to the section "symmetric circuits" and updated every other section in line with changes made.

As our aim is a circuit characterisation of $\FPR$, whose expressive power is
strictly greater than that of $\FPC$, we need to consider gates corresponding to
non-symmetric Boolean functions. In particular, we consider \emph{rank} gates
whose inputs are a matrix and whose result is invariant under row and column
permutations. To lead up to this, we first develop a general framework of
structured Boolean functions. These are functions whose inputs naturally encode
$\rho$-structures and where the output is invariant under the natural symmetries
of such structures. The\emph{matrix-invariant} functions are a natural
partiruclar case. We therefore define symmetric circuits in a general form where
gates can be labelled by \emph{isomorphism-invariant} structured functions.

@@

@@

\subsection{Structured Functions and Symmetry}
Let $X$ be a finite set and $F: \{0,1\}^{X} \rightarrow \{0,1\}$. We call $X$
the \emph{index} of $F$, and denote it by $\ind(F)$. We are often interested in
classes of functions characterised by certain symmetry conditions. Perhaps the
simplest example is the class of symmetric functions, i.e.\ those functions
invariant under the action of the symmetric group on their index. However, in
many cases this symmetry condition may be too strong, or otherwise
inappropriate. For example, suppose $X$ consists of all two element subsets of
some set $V$. We may think of $X$ as containing all potential edges in a graph,
and the input sequence $ \vec{x} \in \{0,1\}^X$ as encoding a graph. In this
case we might require instead that $F$ be invariant under the action of $\sym_V$
on $\vec{x}$. This condition guarantees that $F$ determines a graph property,
and we call such a function \emph{graph-invariant}. Another example of interest,
especially in this paper, is the case where $X := A \times B$, where $A$ and $B$
are non-empty sets, and we think of $\vec{x} \in \{0,1\}^X$ as encoding an
(unordered) matrix. It is natural then to consider the case where $F$ is
invariant under the action of $\sym_A \times \sym_B$, and we say such a function
is \emph{matrix-invariant}. In this subsection we develop a general framework
for dealing with functions that take in structures.

@@

% We should note, however, that symmetry and irreflexivity are
% properties of directed-graph, and as such can be decided by an
% isomorphism-invariant structured function over the graph vocabulary.

@@

\subsection{Symmetric Circuits}
% We define a circuit over a basis of structured functions that are not
% necessarily symmetric. As such, the circuit must include for each gate $g$ an
% index on its children corresponding to the index of the function labelling
% $g$.
In this section we extend the circuit definitions of Anderson and Dawar
\cite{AndersonD17}, introducing a circuit for structures that incorporates an
indexing for each gate in accordance with the Boolean function labelling that
gate. This allows us to define (and evaluate) circuits with gates labelled by
non-symmetric functions.

@@

We now define a \emph{$(\mathbb{B}, \rho)$-circuit}. This is a circuit that
computes a function on structures in the relational vocabulary $\rho$ and where
the gates are labelled by elements of the basis $\mathbb{B}$ consisting of
structured functions. Note that, in the interests of keeping our definitions
from getting excessively complicated, we do not consider circuits computing
functions over multi-sorted vocabularies.

@@

Let $\rho$ be a relational vocabulary, $\mathcal{A}$ a $\rho$-structure over an
$n$-element universe $U$ and $\gamma \in [n]^{\underline{U}}$. Let $\gamma
\mathcal{A}$ be the structure on universe $[n]$ formed by mapping $U$ in
accordance with $\gamma$. The evaluation of a $(\mathbb{B}, \rho)$-circuit $C$
of order $n$ computing a $q$-ary query $Q$ proceeds by recursively evaluating
the gates in the circuit. The evaluation of the gate $g$ for the bijection
$\gamma$ and input $\mathcal{A}$ is denoted by $C[\gamma \mathcal{A}](g)$, and
is given as follows
\begin{enumerate}
  \setlength\itemsep{0mm}
\item If $g$ is a constant gate then it evaluates to the bit given by
  $\Sigma(g)$,
\item if $g$ is a relational gate then $g$ evaluates to true iff $\gamma
  \mathcal{A} \models \Sigma(g)(\Lambda (g))$, and
\item if $g$ is an internal gate such that $\Sigma (g)$ let $L^{\gamma
    \mathcal{A}}(g): \ind(g) \rightarrow \{0,1\}$ be defined by
  $L^{\gamma\mathcal{A}}(g)(x) = C[\gamma \mathcal{A}](L(g)(x))$, for all $x \in
  \ind(g)$. Then $g$ evaluates to true if, and only if, $\Sigma(g)
  (L^{\gamma}_g) = 1$.
\end{enumerate}
We say that $C$ defines the $q$-ary query $Q \subseteq U^q$ under $\gamma$ where
$\vec{a} \in Q$ if, and only if, $C[\gamma \mathcal{A}](\Omega (\gamma \vec{a}))
= 1$.

@@

The following circuit condition is from Anderson and Dawar \cite{AndersonD17}.

@@

It follows that the circuit $C$ of order $n$ is invariant if, and only if, the
evaluation of the circuit for any input $n$-element structure $\mathcal{A}$ does
not depend on the bijection $\gamma$. In general, the evaluation of a circuit
depends on the presentation of the structure (i.e.\ the bijection $\gamma$). If
a family of $(\mathcal{B}, \rho)$-circuits $\mathcal{C}$ is invariant it follows
that the query computed is a $q$-query on $\rho$-structures. Thus if $q = 0$
then $\mathcal{C}$ computes a property of $\rho$-structures. The following lemma
allows us to recast this notion in terms of the language developed in this
paper.

@@

We now define the notion of an automorphism of circuit.

\begin{definition}[Automorphism]
  let $C = \langle G, \Omega, \Sigma, \Lambda, L\rangle$ be a
  $(\mathbb{B},\tau)$-circuit of order $n$ computing a $q$-ary, and where
  $\mathbb{B}$ is a basis of isomorphism-invariant structured functions. Let
  $\sigma \in \sym_n$ and $\pi: G \rightarrow G$ be a bijection such that
  \begin{itemize}
    \setlength\itemsep{0mm}
  \item for all output tuples $x \in [n]^q$, $\pi \Omega (x) = \Omega (\sigma
    x)$,
  \item for all gates $g \in G$, $\Sigma (g) = \Sigma (\pi g)$,
  \item for each relational gate $g \in G$, $\sigma \Lambda (g) = \Lambda (\pi
    g)$, and
  \item For each pair of gates $g, h \in G$ $W(h,g)$ if and only if $W(\pi h,
    \pi g)$ and for each internal gate $g$ we have that $L(\pi g)$ and $ \pi
    \cdot L(g)$ are isomorphism-equivalent.
  \end{itemize}
  We call $\pi$ an \emph{automorphism} of $C$, and we say that $\sigma$
  \emph{extends to an automorphism} $\pi$. The group of automorphisms of $C$ is
  called $\aut (C)$.
\end{definition}

We are particularly interested in circuits that have the property that every
permutation in $\sym_n$ extends to an automorphism of the circuit.

@@

It follows that for any symmetric circuit $C$ of order $n$ there is a
homomorphism $h$ that maps $\sym_n$ to $\aut(C)$ that takes $\sigma \in \sym_n$
to an automorphism that extends it. Suppose $C$ contains no relational gates of
non-zero arity. In that case $C$ computes a constant function. As such, in this
paper we always assume a circuit contains at least one relational gate with
non-zero arity. There there exists a relational gate in $C$ with some element of
$[n]$ appearing in the tuple labelling that gate. By symmetry it follows that
all elements of $[n]$ appear in tuples labelling relational gates in $C$. Thus
no two distinct elements of $\sym_n$ agree on all input gates, and so the
homomorphism $h$ is injective.

If this homomorphism is also surjective then we have that each element of
$\sigma$ extends uniquely to an automorphism of the circuit. Since many of the
technical tools developed in this paper require that circuits have this
property, we restrict our attention to classes of circuits that have
\emph{unique extensions}.

@@

Anderson and Dawar~\cite{AndersonD17} introduce the notion of a \emph{rigid}
circuit, and show that, for circuits defined over a basis of symmetric functions
there is a polynomial-time algorithm that takes as input a symmetric circuit and
outputs an equivalent rigid symmetric circuit. They also show that rigid
circuits have unique extensions, allowing them to conclude that $P$-uniform
families of symmetric circuits over bases of symmetric functions may be assumed
to be rigid (and hence have unique extensions) without a loss of generality.

% It turns out that establishing an exactly analogous argument for general
% circuits is a non-trivial problem.

% In this paper we establish an approximately analogous argument

Anderson and Dawar's argument makes use of the fact that many important circuit
properties are polynomial-time decidable for circuits defined over symmetric
bases. While we should like to construct an analogous argument, we show in
Section~\ref{sec:transparent} that for circuits that include gates computing
non-symmetric functions most of these important circuit properties are at least
as hard to decide as the graph-isomorphism problem. As such, constructing an
analogous translation, as well as establishing the numerous other needed results
whose proofs rely on certain properties being polynomial-time decidable, would
be non-trivial.

In order to proceed we will restrict ourselves to circuits with certain
constraints on how those gates computing non-symmetric functions can be wired
into the circuit. We call these circuits \emph{transparent} circuits. We show in
Sections~\ref{sec:transparent} that transparency is sufficient to allow for the
polynomial-time decidability of many important circuit properties, while still
keeping the circuit model powerful enough for us to establish a translation from
formulas to families of circuits.

Before we can formally define \emph{transparency} we need to define the
\emph{syntactic-equivalence} relation on the gates of a circuit. The intuition
is that two gates in a circuit are syntactically-equivalent if the circuits
underneath them are `hereditary similar'.
% In order to prove this result, Anderson and Dawar introduce the notion of a
% \emph{rigid} circuit, proving that symmetric circuits may be

@@

\begin{definition}
  Let $C := \langle G, \Omega, \Sigma, \Lambda, L \rangle$ be a $(\mathbb{B},
  \rho)$-circuit of order $n$. We recursively define the equivalence relation
  \emph{syntactic-equivalence}, which we denote using the symbol `$\equiv$', on
  $G$ as follows. Suppose $g$ and $h$ are gates in $C$ such that $\Sigma (g) =
  \Sigma(h)$ and either both $g$ and $h$ are output gates and $\Omega^{-1}(g) =
  \Omega^{-1}(h)$ or neither are output gates. Suppose $g$ and $h$ are input
  gates, then $g \equiv h$ if, and only if, both $g$ and $h$ are constant gates
  or both are relational gates and $\Lambda(g) = \Lambda (h)$. Suppose $g$ and
  $h$ are internal gates and and suppose we have defined the
  syntactic-equivalence relation for all gates of depth less than the depth of
  either $g$ or $h$. Then $g \equiv h$ if, and only if, $L(g) /_\equiv$ and
  $L(h) /_\equiv$ are isomorphism equivalent.
\end{definition}

\begin{remark}
  Should I prove these are actually equivalence relations? I use that fact, but
  it's very easy to prove.
\end{remark}

For a circuit $C$ of order $n$ and a gate $g$ we map think of $g$ as computing
the function that maps an input structure $\mathcal{A}$ and a bijection $\gamma$
from the universe of $\mathcal{A}$ to $[n]$ to the evaluation $C[\gamma
\mathcal{A}(g)$. While we would like to be able to identify gates that compute
the same function in this sense, it is not hard to show that deciding this
equivalence relation for a given circuit is $\NP$-hard. We now show that if two
gates are syntactically-equivalent then the functions computed at these two
gates must be equal. We show later that the syntactic-equivalence relation is
polynomial-time decidable for the class of circuits of interest to us in this
paper.

@@

It follows from Lemma~\ref{lem:permutation-extending-syntactic-equivalence} that
the syntactic-equivalence relation of a circuit $C$ constrains the automorphism
group of $C$, and so the orbits and stabiliser groups of the gates in $C$. We
say that a circuit $C$ is \emph{syntactically simple} if $C$ has trivial
syntactic-equivalence classes, i.e. for all gates $g$ and $h$ in $C$ if $g
\equiv h$ then $g = h$. An immediate corollary of
Lemma~\ref{lem:permutation-extending-syntactic-equivalence} is that if a circuit
is syntactically simple then it has unique extensions. We now define a few
properties of circuits in terms of the structure of their syntactic-equivalence
classes.

@@

It is worth noting that if a circuit has injective labels and is syntactically
simple then then it has unique labels. The converse is, in general, false.

@@

It follows from Proposition~\ref{prop:unique-children-unique-extensions} that if
$C$ is of order $n$ then for each $\sigma \in \sym_n$ there exists at most one
$\pi_\sigma \in \aut(C)$ extending $\sigma$. This means that if $C$ is symmetric
for each $\sigma \in \sym_n$ there is always a canonical automorphism
$\pi_\sigma$ extending $\sigma$. However, for a gate $g$ in $C$ there may no
canonical $\lambda \in \aut(g)$ witnessing the validity of the mapping from $g$
to $\pi_\sigma g$, i.e.\ such that $\pi L(g) = L(g') \lambda$. We prove in
Section~\ref{sec:transparent} that in fact, for the class of circuits with
unique children, deciding if there is an isomorphism-equivalence between two
gates is as hard as the graph-isomorphism problem. It follows that a great many
important circuit properties (e.g. the orbit of a gate, symmetry of a circuit,
etc.) for circuits with unique gates are at the very least not obviously
polynomial-time decidable.

In contrast, it is easy to see that if $C$ has unique labels, and $g$ and $g'$
are gates in the circuit $C$, then $L(g)$ and $L(g')$ are isomorphism-equivalent
if, and only if, $\lambda := L(g')^{-1}L(g)$ is in $\aut(g)$. In this case,
rather than having to decide the isomorphism problem we simply need to check if a given
candidate is in fact a valid isomorphism. We prove in
Section~\ref{sec:transparent} that all of the circuit properties of interest in
this paper are polynomial-time decidable for circuits with unique labels.
Moreover, we show in Lemma~\ref{lem:transparent-unique} that a
transparent circuit can be converted into an equivalent circuit with unique
labels in time polynomial in the size of the circuit. Since we are working with
$P$-uniform families of transparent circuits, this fact allows us to assume,
without a loss of generality, that transparent circuits have unique labels (and,
from \ref{prop:unique-children-unique-extensions} so unique extensions).

@@

% Let $\MB$ be the set of all structured functions that are symmetric or
% matrix-invariant.

In many contexts, circuits are defined over a basis of symbols, usually the
standard logical symbols $\land, \neg, \lor$, and perhaps $\maj$ (for majority).
When evaluating a circuit, a gate labelled by a particular symbol is evaluated
using a Boolean function appropriate to the particular symbol. We have instead
defined a basis to be a set of Boolean functions. However, when reasoning about
circuits or dealing with them as inputs to algorithms we often want to identify
a function in the basis with a particular symbol. We now formally define a few
natural bases used in this paper, assigning to each function in these bases an
appropriate symbol.

Let $a \in \nats$, let $\land[a] : \{0,1\}^{[a]} \rightarrow \{0,1\}$ be the
structured function computing the $\AND$ operation on $a$ inputs. Let $\lor[a] :
\{0,1\}^{[a]} \rightarrow \{0,1\}$ be the structured function computing the
$\OR$ operation on $a$ inputs. Let $\maj [a] : \{0,1\}^{[a]} \rightarrow
\{0,1\}$ be the structured function computing the the majority function on $a$
inputs. We abuse notation and let $\neg : \{0,1\} \rightarrow \{0,1\}$ also
denote the structured function computing negation on a single input. In each
case we use the name of the function above as a symbol denoting that function.

Let the \emph{standard basis} $\BS$ be the basis containing $\neg$ and all
functions $\land[a]$ and $\lor[a]$ for every $a \in \nats$. Let the
\emph{majority basis }$\BM := \BS \cup \{\maj[a] : a \in \nats\}$. We say a
basis $\BB$ is \emph{symmetric} if every function in $\mathbb{B}$ is a symmetric
Boolean function.

Let $a, b, r, p \in \nats$, with $p$ prime. Let $\rank^r_p[a,b] : \{0,1\}^{[a]
  \times [b]} \rightarrow \{0,1\}$ be a matrix-invariant structured with
universe $[a] \uplus [b]$, such that $\rank^r_p[a,b](M) = 1$ if, and only if,
the matrix $M \in \{0,1\}^{[a] \times [b]}$ has rank at most $r$ when its
entries are understood as elements of the finite field $\mathbb{F}_p$. We again
identify the function $\rank^{r}_p[a,b]$ with the symbol $\rank^{r}_p[a,b]$.

Let the \emph{rank basis} $\RB := \BM \cup \{\rank^{r}_p[a,b] : a, b, r, p \in
\nats, \text{ $p$ prime}\}$. We say a basis $\BB$ is \emph{matrix-symmetric} if
every function in $\BB$ is symmetric or matrix-symmetric.

>>>>

As our aim is a circuit characterisation of $\FPR$, whose expressive power is
strictly greater than that of $\FPC$, we need to consider gates corresponding to
non-symmetric Boolean functions. In particular, we consider \emph{rank} gates
whose inputs are a matrix and whose result is invariant under row and column
permutations. To lead up to this, we first develop a general framework of
structured Boolean functions. These are functions whose inputs naturally encode
$\tau$-structures and where the output is invariant under the natural symmetries
of such structures. The\emph{matrix-invariant} functions are a natural
partiruclar case. We therefore define symmetric circuits in a general form where
gates can be labelled by \emph{isomorphism-invariant} structured functions.

@@

\begin{drem}
  G: In fact, this difficulty is present throughout the paper.
\end{drem}

@@

\subsection{Structured Functions and Symmetry}
Let $X$ be a finite set and $F: \{0,1\}^{X} \rightarrow \{0,1\}$. We call $X$
the \emph{index} of $F$, and denote it by $\ind(F)$. We are often interested in
classes of functions characterised by certain symmetry conditions. Perhaps the
simplest example is the class of symmetric functions, i.e.\ those functions
invariant under the action of the symmetric group on their index. However, in
many contexts this symmetry condition is inappropriate. For example, suppose $X$
consists of all the two element subsets of some set $V$. So $X$ contains all the
potential edges of a graph with vertex set $V$ and the string $ \vec{x} \in
\{0,1\}^X$ encodes a graph. In this context, it is more natural to require
instead that $F$ be invariant under the action of $\sym_V$ on $\vec{x}$. This
condition guarantees that $F$ is constant on isomorphism classes of graphs with
vertex set $V$, in other words $F$ determines a graph property $F$. We call such
a function \emph{graph-invariant}. Another example of interest, especially in
this paper, is the case where $X := A \times B$, where $A$ and $B$ are non-empty
sets, and we think of $\vec{x} \in \{0,1\}^X$ as encoding an (unordered) matrix.
It is natural then to consider the case where $F$ is invariant under the action
of $\sym_A \times \sym_B$, and we say such a function is
\emph{matrix-invariant}. In this subsection we develop a general framework for
dealing with Boolean functions that take in strings that encode
$\tau$-structures.

@@

% We should note, however, that symmetry and irreflexivity are properties of
% directed-graph, and as such can be decided by an isomorphism-invariant
% structured function over the graph vocabulary.

@@

\subsection{Symmetric Circuits}
% We define a circuit over a basis of structured functions that are not
% necessarily symmetric. As such, the circuit must include for each gate $g$ an
% index on its children corresponding to the index of the function labelling
% $g$.
We now generalise the circuit model of Anderson and Dawar~\cite{AndersonD17} so
as to allow for circuits to be defined over bases that potentially include
non-symmetric (structured) functions. In this model each gate $g$ is not only
associated with an element of the basis, as in the conventional case, but also
with a labelling function. This labelling function maps the input gates of $g$
to an appropriate set of labels (i.e. the index of the structured function
associated with $g$). In concord with this generalisation, we also update the
circuit-related notions discussed by Anderson and Dawar~\cite{AndersonD17},
e.g.\ circuit automorphisms, symmetry, etc. Moreover, we briefly discuss some of
the important complications introduced by our generalisation, and introduce some
of the important tools we will use in later sections to address these
complications.

% in those gates input to each each gate in a circuit is assosiated with both an
% element of the basis \emph{and} a function that labels the input and and the
% circuit also includes a labelling function for each gateThe important addition
% of this generalisation is the inclusion of a labelling function $L$ that
% associates with each gate

% extend the assosiated notion a basis of structured functions. The important to
% be labelled by introducing a circuit for structures that incorporates an
% indexing for each gate in accordance with the Boolean function labelling that
% gate. This allows us to define (and evaluate) circuits with gates labelled by
% non-symmetric functions.

@@

@@

Let $\rho$ be a relational vocabulary, $\mathcal{A}$ be a $\rho$-structure with
universe $U$ of size $n$, and $\gamma \in [n]^{\underline{U}}$. Let $\gamma
\mathcal{A}$ be the structure with universe $[n]$ formed by mapping the elements
of $U$ in accordance with $\gamma$. The evaluation of a $(\mathbb{B},
\rho)$-circuit $C$ of order $n$ computing a $q$-ary query $Q$ proceeds by
recursively evaluating the gates in the circuit. The evaluation of the gate $g$
for the bijection $\gamma$ and input structure $\mathcal{A}$ is denoted by
$C[\gamma \mathcal{A}](g)$, and is given as follows

\begin{myenum}
\item If $g$ is a constant gate then it evaluates to the bit given by
  $\Sigma(g)$,
\item if $g$ is a relational gate then $g$ evaluates to true iff $\gamma
  \mathcal{A} \models \Sigma(g)(\Lambda (g))$, and
\item if $g$ is an internal gate such that $\Sigma (g)$ let $L^{\gamma
    \mathcal{A}}(g): \ind(g) \rightarrow \{0,1\}$ be defined by
  $L^{\gamma\mathcal{A}}(g)(x) = C[\gamma \mathcal{A}](L(g)(x))$, for all $x \in
  \ind(g)$. Then $g$ evaluates to true if, and only if, $\Sigma(g)
  (L^{\gamma}_g) = 1$.
\end{myenum}
We say that $C$ defines the $q$-ary query $Q \subseteq U^q$ under $\gamma$ where
$\vec{a} \in Q$ if, and only if, $C[\gamma \mathcal{A}](\Omega (\gamma \vec{a}))
= 1$.

@@

We notice that in general the output of a circuit $C$ may depend on the
particular encoding of $\mathcal{A}$ as a structure with universe $[n]$, i.e.\
on the chosen bijection $\gamma$. It is natural to consider circuits whose
outputs do do not depend on this choice of encoding. Anderson and
Dawar~\cite{AndersonD17} call such a circuit \emph{invariant}. We have
reproduced their definition below.

@@

If a family of $(\mathcal{B}, \rho)$-circuits $\mathcal{C}$ is invariant it
follows that the query computed is a $q$-query on $\rho$-structures. Thus if $q
= 0$ then $\mathcal{C}$ computes a property of $\rho$-structures. The following
lemma allows us to recast this notion in terms of the language developed in this
paper.

@@

We now define an automorphism of a circuit, generalising the definition
introduced by Anderson and Dawar. The definition is similar, but adds the
requirement that if a gate $g$ is mapped to $g'$, then children of $g$ must be
mapped to the children of $g'$ via some appropriate isomorphism of the structure
associated with $g$.

\begin{definition}[Automorphism]
  let $C = \langle G, \Omega, \Sigma, \Lambda, L\rangle$ be a
  $(\mathbb{B},\tau)$-circuit of order $n$ computing a $q$-ary, and where
  $\mathbb{B}$ is a basis of isomorphism-invariant structured functions. Let
  $\sigma \in \sym_n$ and $\pi: G \rightarrow G$ be a bijection such that
  \begin{myitemize}
  \item for all output tuples $x \in [n]^q$, $\pi \Omega (x) = \Omega (\sigma
    x)$,
  \item for all gates $g \in G$, $\Sigma (g) = \Sigma (\pi g)$,
  \item for each relational gate $g \in G$, $\sigma \Lambda (g) = \Lambda (\pi
    g)$, and
  \item For each pair of gates $g, h \in G$ $W(h,g)$ if and only if $W(\pi h,
    \pi g)$ and for each internal gate $g$ we have that $L(\pi g)$ and $ \pi
    \cdot L(g)$ are isomorphism-equivalent.
  \end{myitemize}
  We call $\pi$ an \emph{automorphism} of $C$, and we say that $\sigma$
  \emph{extends to an automorphism} $\pi$. The group of automorphisms of $C$ is
  called $\aut (C)$.
\end{definition}

We are particularly interested in circuits that have the property that
\emph{every} permutation in $\sym_n$ extends to an automorphism of the circuit.

@@

It follows that for any symmetric circuit $C$ of order $n$ there is a
homomorphism $h$ that maps $\sym_n$ to $\aut(C)$ such that if $\sigma \in
\sym_n$ then $h(\sigma)$ is an an automorphism extending $\sigma$. Suppose $C$
does not contain a relational gate labelled by a relation symbol with non-zero
arity. In that case $C$ computes a constant function. As such, in this paper we
always assume a circuit contains at least one relational gate with non-zero
arity. As such by assumption there exists a relational gate in $C$ such that
some element of $[n]$ appears in the tuple labelling that gate. By symmetry it
follows that every elements of $[n]$ appears in a tuple labelling a relational
gate in $C$. It follows that no two distinct elements of $\sym_n$ agree on all
input gates, and so the homomorphism $h$ is injective.

If this homomorphism is also surjective then we have that each element of
$\sigma$ extends uniquely to an automorphism of the circuit. In this case we say
that a circuit has \emph{unique extensions}.

@@

It is worth noting that many of the important technical tools needed in this
paper are only applicable if the circuit under consideration has unique
extensions. In order to handle this technicality Anderson and
Dawar~\cite{AndersonD17} introduce the notion of a \emph{rigid} circuit and show
that, for circuits defined over a basis of symmetric functions there is a
polynomial-time algorithm that takes as input a symmetric circuit and outputs an
equivalent rigid symmetric circuit. They also show that a great many properties
of rigid circuits can be decided in polynomial time, a necessary step in their
argument, and, importantly, that that rigid circuits have unique extensions.
This allows them to restrict their attention to $P$-uniform families of
symmetric circuits (over bases of symmetric functions) \emph{with unique
  extensions}, without a loss of generality.

In order to make use of these technical tools, as well ensure the
polynomial-time decidability of many important circuit properties, we should
like to be able to construct a normal form analogous to rigidity and present a
polynomial time algorithm that transforms a circuit into an equivalent circuit
of this form. It is at this point that we arrive at the first complication
introduced by our generalisation. Anderson and Dawar's polynomial-time
translation makes indispensable of the polynomial-time decidability of many
important circuit properties for circuits defined over symmetric bases. However,
for the more general circuits discussed here, it is not known if even the most
basic circuit properties are polynomial-time decidable. Indeed, we show in
Section~\ref{sec:transparent} that for circuits that include gates computing
non-symmetric functions all of the relevant circuit properties of interest are
at least as hard to decide as the graph-isomorphism problem. As such,
constructing an analogous argument, as well as establishing the numerous other
crucial results whose proofs rely on the polynomial-time decidability of various
circuit properties, would be beyond the scope of this paper.

In order to proceed we explicitly restrict our attention to a particular class
of circuits characterised by a restriction on the children of gates labelled by
non-symmetric functions. We say such circuits are \emph{transparent}. We show in
Section~\ref{sec:transparent} that all of the circuit properties of interest are
polynomial-time decidabile for transparent circuits, and we use these results to
define a polynomial-time transformation from transparent circuits to equivalent
circuits with unique extensions. Importantly, while the class of transparency
condition is a constraint sufficient to enable our analysis, we show in
Section~\ref{sec:formulas-to-circuits}, that the class of transparent circuits
is still rich enough to capture $\FPR$.

Before we can formally define \emph{transparency} we need to define the
\emph{syntactic-equivalence} relation on the gates of a circuit. The intuition
is that two gates in a circuit are syntactically-equivalent if the circuits
underneath them are `hereditary equivalent', i.e.\ the two circuits underneath
these gates are really just copies of one another.
% In order to prove this result, Anderson and Dawar introduce the notion of a
% \emph{rigid} circuit, proving that symmetric circuits may be

@@

\begin{definition}
  Let $C := \langle G, \Omega, \Sigma, \Lambda, L \rangle$ be a $(\mathbb{B},
  \rho)$-circuit of order $n$. We recursively define the equivalence relation
  \emph{syntactic-equivalence}, which we denote using the symbol `$\equiv$', on
  $G$ as follows. Suppose $g$ and $h$ are gates in $C$ such that $\Sigma (g) =
  \Sigma(h)$ and either both $g$ and $h$ are output gates and $\Omega^{-1}(g) =
  \Omega^{-1}(h)$ or neither are output gates. Suppose $g$ and $h$ are input
  gates, then $g \equiv h$ if, and only if, both $g$ and $h$ are constant gates
  or both are relational gates and $\Lambda(g) = \Lambda (h)$. Suppose $g$ and
  $h$ are internal gates and and suppose we have defined the
  syntactic-equivalence relation for all gates of depth less than the depth of
  either $g$ or $h$. Then $g \equiv h$ if, and only if, $L(g) /_\equiv$ and
  $L(h) /_\equiv$ are isomorphism-equivalent.
\end{definition}

\begin{drem}
  Should I prove these are actually equivalence relations? I use that fact, but
  it's very easy to prove.
\end{drem}

For a circuit $C$ of order $n$ and a gate $g$ we map think of $g$ as computing
the function that maps an input structure $\mathcal{A}$ and a bijection $\gamma$
from the universe of $\mathcal{A}$ to $[n]$ to the evaluation $C[\gamma
\mathcal{A}(g)$. While we would like to be able to identify gates that compute
the same function in this sense, it is not hard to show that deciding this
equivalence relation for a given circuit is $\NP$-hard. We now show that if two
gates are syntactically-equivalent then the functions computed at these two
gates must be equal. We show later that the syntactic-equivalence relation is
polynomial-time decidable for the class of circuits of interest to us in this
paper. In this sense we shall treat syntactic-equivalence as a tractable
refinement of the $\NP$-complete relation.

@@

It follows from Lemma~\ref{lem:permutation-extending-syntactic-equivalence} that
the syntactic-equivalence relation of a circuit $C$ constrains the automorphism
group of $C$, and so the orbits and stabiliser groups of the gates in $C$. We
say that a circuit $C$ is \emph{reduced} if $C$ has trivial
syntactic-equivalence classes, i.e. for all gates $g$ and $h$ in $C$ if $g
\equiv h$ then $g = h$. An immediate corollary of
Lemma~\ref{lem:permutation-extending-syntactic-equivalence}, and en example of
this intuitive understanding of the result, is that if a circuit is reduced then
it has unique extensions. We now define \emph{transparency} and other properties
of circuits in terms of the structure of their syntactic-equivalence classes.

@@

It is worth noting that if a circuit has injective labels and is reduced then
then it has unique labels. The converse is, in general, false.

@@

We have from Proposition~\ref{prop:unique-children-unique-extensions} that
circuits with unique gates have unique extensions, but, in order to prove many
of the results we will need later, we not only need that each automorphism is
uniquely defined by the permutation it extends, but also that the isomorphisms
that witness the isomorphism-equivalences, and so witness the validity of the
automorphism, are also uniquely defined (or, at least very least, well
constrained) by the permutation $\sigma$. Without a sufficient constraint on the
choice of isomorphism, testing if two gates have isomorphism-equivalent
labellings (a requirement for computing the automorphism extending a
permutation) requires solving some version of the isomorphism problem for
$\tau$-structures, where $\tau$ is the vocabulary of the gates in question.
Indeed, we prove in Section~\ref{sec:transparent}, that for the class of
circuits with unique children deciding if there is an isomorphism-equivalence
between two gates is as hard as the graph-isomorphism problem. It follows that a
great many important circuit properties (e.g. the orbit of a gate, symmetry of a
circuit, etc.) for circuits with unique gates are at the very least not
obviously polynomial-time decidable. As such, circuits with unique children,
although satisfying the requirement of having unique extensions, are
insufficient for our purposes.

\begin{drem}
  G: I think the above paragraph may be getting a bit too far into the weeds.
\end{drem}

In order to ensure a canonical choice of isomorphism we instead consider
circuits with unique labels. It is easy to see that if $C$ has unique labels
then $L(g)$ and $L(g')$ are isomorphism-equivalent if, and only if, $\lambda :=
L(g')^{-1}L(g)$ is an isomorphism. In this case, rather than having to decide
the isomorphism problem, we can decide isomorphism-equivalences by checking if a
given candidate function is a valid isomorphism. We prove in
Section~\ref{sec:transparent} that all of the circuit properties of interest in
this paper are polynomial-time decidable for circuits with unique labels.
Moreover, we show in Lemma~\ref{lem:transparent-unique} that a transparent
circuit can be transformed into an equivalent circuit with unique labels in time
polynomial in the size of the circuit. Since we are working with $P$-uniform
families of transparent circuits, this fact allows us to assume, without a loss
of generality, that transparent circuits have unique labels and so , from
\ref{prop:unique-children-unique-extensions}, unique extensions).

@@

% Let $\MB$ be the set of all structured functions that are symmetric or
% matrix-invariant.
We now discuss a few Boolean basis of interest in this paper and state the main
result. In most cases (e.g.\ \cite{}) a Boolean basis is taken to be a finite
sequence of symmetric Boolean functions. A binary encoding of a circuit defined
over such a basis is given by assigning to each element of the basis a unique
symbol, and then encoding the circuit in the obvious manner. Since there are
only a finite number of basis elements, it follows that if each basis element
can be computed in polynomial time then the mapping that takes a symbol $s$ and
a string $\vec{x}$ to the value of the function denoted by $s$ for input
$\vec{x}$ can be decided in polynomial-time. We call this function the
\emph{gate evaluation function}, and note that its polynomial-time decidability
is required for proving the polynomial-time evaluation of the circuit. However,
in order to allow for the inclusion of structured (possibly non-symmetric)
functions in the basis, we have instead defined a basis to be a (possibly
infinite) set of finite Boolean functions. In this case the gate evaluation
function may not even be decidable, let alone polynomial-time decidable.

\begin{definition}
  Let $\BB$ be a basis. Let $S$ be a set of symbols and let $F : \BB \rightarrow
  S$ be a bijection. We say $\BB$ is \emph{effective} if the function $E$ that
  maps a symbol $s$ and an input $x \in \dom(F(s))$ to the value
  $(F^{-1}(s))(x)$ is decidable. We say $\BB$ is \emph{$\PT$-effective} if $E$
  is polynomial-time decidable.
\end{definition}

It is easy to see that if $\BB$ is a $\PT$-effective basis then we can evaluate
a circuit over $\BB$ in time polynomial in the size of the circuit.

Let $n \in \nats$, we have that $\AND[n] : \{0,1\}^{[n]} \rightarrow \{0,1\}$ is
the restriction (as a structured function) of the $\AND$ function to inputs of
size $n$. Similarly, we have that $\OR[n]$, $\NAND[n]$ and $\MAJ[n]$ are the
restrictions of these respective functions to strings of length $n$. We let
$\NOT: \{0,1\} \ra \{0,1\}$ be the not function, and we often identify $\NOT$
with $\NAND[1]$. In each case we use the name of the function as the symbol
denoting that function.

As such, the standard basis $\BS$ consists of the functions $\NAND[n]$,
$\AND[n]$, and $\OR[n]$ for every $n \in \nats$, and the majority basis is $\BM
:= \BS \cup \{\MAJ[n] : n \in \nats\}$. We often use $\NAND[1]$ and $\NOT$
interchangeably. We say a basis $\BB$ is \emph{symmetric} if every function in
$\BB$ is a symmetric Boolean function.

Let $a, b, r, p \in \nats$, with $p$ prime. Let $\RANK^r_p[a,b] : \{0,1\}^{[a]
  \times [b]} \rightarrow \{0,1\}$ be a matrix-invariant structured function
with universe $[a] \uplus [b]$, such that $\RANK^r_p[a,b](M) = 1$ if, and only
if, the matrix $M \in \{0,1\}^{[a] \times [b]}$ has rank at most $r$ over $\ff_p$
when the entries of $M$ are interpreted as elements of $\mathbb{F}_p$. We
identify the function $\RANK^{r}_p[a,b]$ with the symbol $\RANK^{r}_p[a,b]$.
